What Makes Deep Network Representations Good for Visual Recognition?

### 2015-04-03 paper_for_review.tex
Corrections throughout paper.

%%%%%%%%% TITLE
\title{What Makes Deep Network's Representations Good for Visual Recognition?}

@@

Representations of artificial neurons in deep neural networks have been demonstrated highly effective and robust in various visual recognition tasks \cite{krizhevsky2012imagenet, sermanet2013overfeat, donahue2014decaf}.
Nevertheless, our understanding of them arguably remains limited.
In this paper, we address two important and practical questions: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
Using methods which support characterizing both first-order and second-order properties of scalar and vector representations, we conducted a large-scale experiment, with 200 randomly generated shallow and deep networks, or totally 6,400 neurons characterized, and the results analyzed through a series of measures designed to quantify the differences in representations. 
We conclude that deep representations are quantitatively better than shallow representations, and overall 71\% of the variance in deep networks' performances can be explained with our measures, both of which are confirmed using statistical tests for the first time.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.40\textwidth]{Figs/e_fig7s_compact-crop.pdf} 
\end{center}
\caption{{\bf What makes a deep network good?}
In existing literatures, researchers mostly tried to come up with one (or few) ``best'' network architecture (\ie hyperparameter set, either through manual tuning or automated hyperparameter search) and then benchmarked the network's performances.
However, as exemplified in the figure, when considering a larger population of networks \cite{cox2011beyond}, it is not fully clear to us why deep (\eg two-level) networks in practice outperform shallow (\eg one-level) networks, and why certain networks are better than the others, given their broad distribution of performances on LFW-a face recognition \cite{LFWTech, wolf2011effective}.
%However, it is not fully clear that in practice (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
%As exemplified in the figure, deep (\ie two-level) networks not only on average outperform shallow (\ie one-level) networks \cite{cox2011beyond}, but also exhibit a wide range of performance differences on LFW-a face recognition \cite{LFWTech, wolf2011effective}.
} % (\eg optimal stimulus). %and (sometimes) characterized its properties
\label{fig:perf}
\end{figure}

Although deep neural networks have been successfully pushing the boundaries of computer vision by breaking the records of ImageNet Challenges (ILSVRC) year after year recently \cite{krizhevsky2012imagenet, sermanet2013overfeat, szegedy2014going}, our understanding about the properties of these networks arguably remains limited, since we could not yet fully answer the question ``what makes a deep network good?''.
As demonstrated in Fig.~\ref{fig:perf}, we summarized two important subquestions to be investigated: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
Specifically, from the point of view that successful networks are forming ``better'' representations of visual inputs that are more separable to the final-stage linear classifiers, we want to identify the crucial properties that are required in forming (or learning) better representations.

@@

%\newcommand{\ivec}{\mathrm{vec}^{-1}}
%$f({\bf{x}}) = \left\| {\bf{K}} \otimes \ivec\left({\bf{x}}\right) \right\|_{F}$
\newcommand{\expstimdim}{For simplicity of formulas, a stimulus being 2-dimensional (\ie ${\bf{x}} \in \mathbb{R}^{\sqrt{N} \times \sqrt{N}}$) or vectorized (\ie ${\bf{x}} \in \mathbb{R}^N$) are used interchangeably.}

%first order, second order, flexibility
{\bf Existing works.} 
Given an artificial neural network $f$, the studying of how $f$ represents a high dimensional stimulus\footnote{\expstimdim} ${\bf{x}} \in \mathbb{R}^N$ as a representation 
${\bf{r}} = f\left( {\bf{x}} \right)$ in existing literatures mainly focused on the first-order (\ie linear) characterization of $f$---the analysis of optimal stimuli, stimuli which lead to the strongest responses of given neurons.
For single-level convolutional networks using squared pooling, \ie $f({\bf{x}}) = \left\| {\bf{K}} \otimes {\bf{x}} \right\|_{F}$, optimal stimulus was shown to be analytically approximable \cite{saxe2011random} using a Gabor-like filter, whether the convolution kernel $\bf{K}$ is structural or random.
For multi-level convolutional networks, optimal stimuli of neurons from various depths were numerically approximated \cite{ngiam2010tiled, le2012building, zeiler2014visualizing, simonyan2013deep} to qualitatively show neurons in deeper layers are tuned to more complex visual patterns.
The generalized concept of optimal stimulus in considering groups of neurons can also be adopted to study the representations by inverting them \cite{mahendran2014understanding}.

In addition to the first-order characterization, the analyses of invariance and selectivity properties of neurons (\eg changes in stimuli which given neurons are least and most sensitive to, how small or large the changes are, \etc) was adopted in a few works, which can be viewed as the second-order characterization of $f$, and parametric deformation in stimuli was the most commonly used setup.
For example, translation, rotation, scaling, \etc in stimulus were used in \cite{goodfellow2009measuring, zeiler2014visualizing} to test the invariance and selectivity of neurons.
Similar deformations were also adopted in \cite{lenc2014understanding}, but extended to analyze more advanced properties (\ie equivariance and equivalence) of groups of neurons.
More generalized setups that, \eg, utilize local quadratic information (\ie the Hessian) or ``global'' numerical searches in the stimulus space, even though do exist, were less explored in the context of artificial neural network studies.
For quadratic networks, \ie $f\left({\bf{x}}\right) = \frac{1}{2}{\bf{x}}^{T}{\bf{Qx}}+{\bf{L}}^{T}{\bf{x}}+c$, through eigendecomposing the quadratic term $\bf{Q}$ (\ie the Hessian), local invariance and selectivity directions can be analytically derived \cite{berkes2006analysis}, which correspond to eigenvectors of the least and most negative eigenvalues.
This method was also generalized onto multi-layer convolutional networks through numerically estimating the Hessian \cite{ngiam2010tiled}.
For multi-layer networks, characterization of the ``global'' invariance of single neurons can also be achieved through numerically searching for non-local (\ie farther to the optimal stimuli) solutions \cite{erhan2010understanding}.

@@

\section{Methods}

\begin{figure}
\begin{center}
\includegraphics[width=0.35\textwidth]{Figs/methods.pdf} 
\end{center}
\caption{{\bf Representation characterization methods.}
On the spherical (\ie constant energy) constraint of the $N$ dimensional stimulus space, similar to previous works, the first-order property (\ie optimal stimulus $\hat{\bf{x}}$, as shown in the figure's center) of a representation $\bf{r}$ is first iteratively searched (as the evolution of light gray eclipses shown in the figure) and analyzed.
However, unlike most previous works, the second-order properties (\ie invariant path $\lbrace \bf{x}^{+}_{\delta} \rbrace$ as dashed red curve and selectivity path $\lbrace \bf{x}^{-}_{\delta} \rbrace$ as dashed blue curve) at distances $0.1\pi$ to $0.5\pi$ from the optimal stimulus are further searched and analyzed.
Additionally, these methods are capable of characterizing both scalar and vector representations, where the ``fitnesses'' to be optimized correspond to Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1} and Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}, respectively.
In this example, the scalar representation (\ie response) of the given neuron is tuned to a $45^{\circ}$ Gabor filter and invariant to phase changes, but selective to orientation changes.
}
\label{fig:methods}
\end{figure}

@@

To study the representations deep inside an artificial neural network, methods adopted in this work include: (1) first-order (\ie linear) characterization, or search and analysis of optimal stimuli, and (2) second-order (\ie quadratic) characterization, or search and analysis of invariance and selectivity paths (\ie eigenvectors of the largest and smallest eigenvalues), which are generalized and improved from previous works.
$\bf{r}$ is defined as a scalar representation (\ie $\bf{r} \in \mathbb{R}$) when considering single neurons and a vector representation (\ie ${\bf{r}} \in \mathbb{R}^C$) when considering groups of neurons (with group size $C$).
Our methods support characterizing both scalar (using Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1}, though Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2} can be used as well) and vector (using Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}) representations in a unified way.
The search phases defined via these equations will be introduced in the following subsections, as the analyses of search results using representation measures will be covered in Sec.~\ref{sec:exp} when applied.

@@

While no specific knowledge of the network itself is assumed, we do restrict the space of stimuli considered with a constant energy constraint $\left\| \bf{x} \right\| = E$, where $E$ is set to $1$ in all mathematical derivations for simplicity, but valued as the average energy of task-related images in our experiments.
Since neurons can often be modulated by stimulus contrasts, which is less interesting when considering pattern selectivities, limiting the stimulus space in this manner effectively avoids degenerate solutions that simply maximize stimulus contrasts, and reduces the range of possible stimuli to consider.

@@

In this work, optimal stimulus is defined and numerically derived as 
\begin{align}
\hat{\bf{x}} &= \underset{\bf{x}} {\arg\max} f\left(\bf{x}\right) \label{eq:O1} \\
&= \underset{\bf{x}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}\right)-f\left(\tilde{\bf{x}}\right)\right\|} \right); \label{eq:O2}
\end{align}
subject to the spherical constraint $\left\| \bf{x} \right\| = 1$. 
As visualized in Fig.~\ref{fig:methods}, this process involves an iterative search on the non-convex fitness landscape, which in fact can be handled efficiently by most of the modern numerical solvers, especially when analytical derivatives (\ie backpropagation gradients) are available.
In this work, we adopted the CMA-ES solver \cite{hansen2001completely}, which though do not rely on analytical gradients and instead stochastically estimates numerical derivatives (\ie inverse Hessian), {worked} reasonably well in our experiments.
When considering a vector representation (using Eq.~\ref{eq:O2}), this process is equivalent to optimizing the response of an ``imaginary neuron'' tuned to a reference stimulus $\tilde{\bf{x}}$.

@@

With respect to the optimal stimulus $\hat{\bf{x}}$, the searches of invariance and selectivity paths, which consist of sets of invariant stimuli $\left\lbrace \bf{x}^{+}_{\delta} \right\rbrace$ and selective stimuli $\left\lbrace \bf{x}^{-}_{\delta} \right\rbrace$ respectively, are formulated as
\begin{align}
\bf{x}^{+}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\max} f\left(\bf{x}_{\delta}\right) \label{eq:I1} \\
&= \underset{\bf{x}_{\delta}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:I2} \\
\bf{x}^{-}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\min} f\left(\bf{x}_{\delta}\right) \label{eq:S1} \\
&= \underset{\bf{x}_{\delta}}{\arg\min} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:S2}
\end{align}
in which $0 < \delta \le \frac{\pi}{2}$, subject to $\left\| \bf{x}_{\delta} \right\| = 1$ and $\langle \bf{x}_{\delta} , \hat{\bf{x}} \rangle = \cos\left(\delta\right)$. 
As visualized in Fig.~\ref{fig:methods}, the invariance path characterizes the $N$ dimensional curve that leads toward (or ideally, maintains at) fitness as high as possible while moving away from the optimal stimulus, and the selectivity path characterizes the curve that leads toward fitness as low as possible while moving away.
The search process is implemented as multiple runs of maximization and minimization on discretized $\delta \in \left\lbrace 0.1\pi, 0.2\pi, 0.3\pi, 0.4\pi, 0.5\pi\right\rbrace$ as the circular distance constraints shown in Fig.~\ref{fig:methods}, where each run is initialized with the result from previous run (and the $0.1\pi$ run directly with optimal stimulus $\hat{\bf{x}}$) to increase the path continuity and searching speed\footnote{\expdiff}.
In this work, the distance constraint $\delta$ only goes up to $\frac{\pi}{2}$, since on the $N$ dimensional sphere, stimuli fall in $\frac{\pi}{2} < \delta \le \pi$ are simply ``negatives'' of those in $0 < \delta \le \frac{\pi}{2}$, which are of less uniqueness and interest in general. 
The same process can also be performed with respect to a reference stimulus $\tilde{\bf{x}}$, especially for vector representations, where the invariance and selectivity of the ``imaginary neuron'' are again characterized equivalently.

@@

For our experiments, a simplified variety of deep convolutional neural networks---the Spatio-Temporal Hierarchical Object Representation (STHOR) model \cite{pinto2009high, sthor}---{was} adopted.
Although like other deep networks, it also consists of the standard cascade of convolution, nonlinear activation, pooling layers, \etc (which together define a single ``level'' in this work), its convolution kernel weights are however randomly assigned rather than trained.
Though not matching the performances of deep networks trained on massive quantities of data \cite{krizhevsky2012imagenet}, it can in fact be surprisingly competitive for a variety of tasks, especially in the regime where large quantities of training samples are not easily available \cite{pinto2009high, cox2011beyond, viglarge}.
Most importantly, since it does not require training (except for the linear SVM on top of the representations extracted), a large number of networks with diverse hyperparameters can be rapidly generated and compared, which is particularly useful in this study where relative, instead of absolute task performances, are of the main interest. 
As one might concern that comparing randomly weighted networks will not tell much about representations, since they might as well be random, our results also suggest otherwise.

\newcommand{\expsettings}{In other words, 32 channels of filters in top levels' convolution layers, a simple setting with reasonable performances.
Following the terminology in recent literatures, shallow and deep neurons corresponded to \texttt{pool1} and \texttt{pool2} layers respectively.
Stimulus dimensionalities of the shallow and deep neurons {were} $N=121$ and $441$ (\ie spatially overlapping $11\times11$ and $21\times21$ receptive fields). % respectively
Other minor changes included: nonlinear activations all simplified to \emph{ReLU} \cite{krizhevsky2012imagenet} mode and normalizations all subtractive mode.
Overall, the architectures were more similar to those in \cite{simonyan2014very}, except pooling operations can be \emph{average}, \emph{squared}, or \emph{max-like} in our case.
}

@@

In the experiments, 100 shallow (\ie one-level) networks and 100 deep (\ie two-level) networks {were} randomly generated (except all with 32 top-layer neurons\footnote{\expsettings}) and tested to see how their representations differ with networks' depths and affect networks' performances on face pair matching\footnote{\exptask} tasks against the LFW-a dataset \cite{LFWTech, wolf2011effective} as shown in Fig.~\ref{fig:perf}. %, using 10-fold cross validation on view 2.
The fitness landscapes of shallow and deep representations in our case are both non-convex, therefore there is no guarantee that optimality can be achieved through any optimization method.
However, by carefully adjusting the numerical solver and monitoring the quality of solutions, we argue that informative local optima can still be discovered and used to characterize the representations.
Specifically, all numerical optimizations {were} executed multiple times from different random initial stimuli, for increasing quality of the numerical solutions (of optimal stimuli and invariance and selectivity paths, with 2 runs executed and the better kept), and at the same time, for providing statistical samplings of the solution spaces (of encoding specificities and invariance and selectivity subspace properties, as detailed later).

@@

{\bf Optimal stimulus spectral complexity.} Figure \ref{fig:allrep}A and \ref{fig:allrep}E illustrate the trajectories of optimal stimulus searches of 32 top-layer neurons from the best performing shallow network and 32 from the best performing deep network respectively.
To further increase the searching speed, the initial stimulus {was} selected from 1000 ${1}/{f^{\alpha}}$ random stimuli with the best fitness, where $\alpha \in \left\lbrace -4,-3,-2,-1,0 \right\rbrace$, without overly sacrificing the nature of random initialization.
From the results, first, the response landscapes of deep neurons are noticeably more complex than those of shallow neurons, as a large fraction of searching trajectories for deep neurons actually go down then up before reaching the (local) optima, which also suggests the numerical solver can handle non-convexity reasonably well.
Second, the optimal stimuli of deep neurons are also relatively more complex (\ie consisting of more textural components) than those of shallow neurons (which are more uni-textured or Gabor-like).
To quantify such differences, the spectral complexity measure, which is estimated through the $L^{1}$ norm of the (2 dimensional) Fourier power spectrum of optimal stimulus, \ie $\left\| \mathcal{F}(\hat{\bf{x}}) \right\|_{1}$, is adopted, where higher value suggests higher non-sparsity (\ie more spectral components required to represent the visual stimuli).

@@

{\bf Invariance subspace capacity.} To further characterize the invariance properties, multiple runs of invariance path searches {were} performed and the results are visualized in Fig.~\ref{fig:allrep}D and \ref{fig:allrep}H.
The diversity of the ``invariance plateau'' (\eg the central high fitness region in Fig.~\ref{fig:methods}) constitution of deep neurons is also more apparent than that of shallow neurons.
Compared to invariance path potential which is designed to characterize the best invariance path even if only one of such exists, we further adopted the invariance subspace capacity measure to estimates the ``dimensionality'' (\ie how diverse different paths can be) of the linear subspace formed by multiple path search results via the nuclear norm of concatenation of $n$ results $\left\|\left[{\bf{x}}_{\delta,1},\dots,{\bf{x}}_{\delta,n}\right]\right\|_{*}$ with $n=20$ and $\delta=0.1\pi$.

@@

\newcommand{\expregular}{Equation \ref{eq:O2} though describes the essence of our approach, was not able to produce analyzable numerical results alone, due to the fact that deep representations can be highly invariant to noises that do not resemble natural images.
Implementation-wise, it was further regularized with the energy of Laplacian filtering $\left\| \mathcal{L} \left( \bf{x} \right) \right\|$ to enforce smoothness and similarity to natural images.
Similar concept was also adopted in \cite{mahendran2014understanding} independently.
}  %, with regularization constant decided via optimizing withheld data.

>>>>

%%%%%%%%% TITLE
\title{What Makes Deep Network Representations Good for Visual Recognition?}

@@

Representations of artificial neurons in deep neural networks have been demonstrated to be highly effective and robust for various visual recognition tasks~\cite{krizhevsky2012imagenet, sermanet2013overfeat, donahue2014decaf}.
Nevertheless, our understanding of them arguably remains limited.
In this paper, we address two important and practical questions: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others. Using methods which support characterizing both first-order and second-order properties of scalar and vector representations, we considered the case study of face pair matching and conducted a large-scale experiment. This experiment examined 200 randomly generated shallow and deep networks, with the results analyzed through a series of measures designed to quantify the differences in representations. 
We conclude though this principled analysis that deep representations are quantitatively better than shallow representations, and overall 71\% of the variance in the performance of deep networks can be explained with our measures, both of which are confirmed using statistical tests for the first time.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.40\textwidth]{Figs/e_fig7s_compact-crop.pdf} 
\end{center}
\caption{{\bf What makes a deep network good?}
In existing work, researchers have mostly tried to produce one ``best'' network architecture (\ie a hyperparameter set, either through manual tuning or automated hyperparameter search), which is subsequently evaluated on a particular benchmark. However, as exemplified in this figure, when considering a larger population of networks \cite{cox2011beyond}, it is not fully clear to us why in practice deep networks (\eg two or more levels) outperform shallow networks (\eg one-level), and why certain networks are better than the others, given their broad distribution of performance on a task like face pair matching for the LFW-a data set (shown above)~\cite{LFWTech, wolf2011effective}.
%However, it is not fully clear that in practice (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than the others.
%As exemplified in the figure, deep (\ie two-level) networks not only on average outperform shallow (\ie one-level) networks \cite{cox2011beyond}, but also exhibit a wide range of performance differences on LFW-a face recognition \cite{LFWTech, wolf2011effective}.
} % (\eg optimal stimulus). %and (sometimes) characterized its properties
\label{fig:perf}
\end{figure}

Although deep neural networks have been successfully pushing the boundaries of computer vision by breaking the records of ImageNet Challenges (ILSVRC) year after year~\cite{krizhevsky2012imagenet, sermanet2013overfeat, szegedy2014going}, our understanding of the properties of these networks arguably remains limited, since we cannot yet fully answer the question ``what makes a deep network good?'' In this paper, we investigate two important facets of this question: (1) why deep networks are better than shallow networks, and (2) among networks of the same depth, why certain networks are better than others (Fig.~\ref{fig:perf}). Specifically, from the point of view that successful networks are forming ``better'' representations of visual inputs that are more separable to the final-stage linear classifiers, we want to identify the crucial properties that are required for forming (or learning) better representations.

@@

%\newcommand{\ivec}{\mathrm{vec}^{-1}}
%$f({\bf{x}}) = \left\| {\bf{K}} \otimes \ivec\left({\bf{x}}\right) \right\|_{F}$
\newcommand{\expstimdim}{For simplicity, a stimulus being 2-dimensional (\ie ${\bf{x}} \in \mathbb{R}^{\sqrt{N} \times \sqrt{N}}$) or vectorized (\ie ${\bf{x}} \in \mathbb{R}^N$) are used interchangeably.}

%first order, second order, flexibility
{\bf Existing work.} 
Given an artificial neural network $f$, the study of how $f$ represents a high dimensional stimulus\footnote{\expstimdim} ${\bf{x}} \in \mathbb{R}^N$ as a representation 
${\bf{r}} = f\left( {\bf{x}} \right)$ in existing work mainly focused on the first-order (\ie linear) characterization of $f$ --- the analysis of optimal stimuli, stimuli which lead to the strongest responses of given neurons.
For single-level convolutional networks using squared pooling, \ie $f({\bf{x}}) = \left\| {\bf{K}} \otimes {\bf{x}} \right\|_{F}$, the optimal stimulus was shown to be analytically approximable \cite{saxe2011random} using a Gabor-like filter, whether the convolution kernel $\bf{K}$ is structural or random.
For multi-level convolutional networks, the optimal stimuli of neurons from various depths were numerically approximated \cite{ngiam2010tiled, le2012building, zeiler2014visualizing, simonyan2013deep} to qualitatively show neurons in deeper layers are tuned to more complex visual patterns.
The generalized concept of optimal stimulus in considering groups of neurons can also be adopted to study the representations by inverting them \cite{mahendran2014understanding}.

In addition to the first-order characterization, the analysis of invariance and selectivity properties of neurons (\eg changes in stimuli which given neurons are least and most sensitive to, how small or large the changes are, \etc) was adopted in a few works. This can be viewed as the second-order characterization of $f$, and parametric deformation in stimuli was the most commonly used setup.
For example, translation, rotation, scaling, \etc of the stimulus were used in~\cite{goodfellow2009measuring, zeiler2014visualizing} to test the invariance and selectivity of neurons.
Similar deformations were also adopted in \cite{lenc2014understanding}, but extended to analyze more advanced properties (\ie equivariance and equivalence) of groups of neurons.
More generalized setups that, \eg, utilize local quadratic information (\ie the Hessian) or ``global'' numerical searches in the stimulus space, even though they do exist, are less explored in the context of artificial neural network studies.
For quadratic networks, \ie $f\left({\bf{x}}\right) = \frac{1}{2}{\bf{x}}^{T}{\bf{Qx}}+{\bf{L}}^{T}{\bf{x}}+c$, through eigendecomposing the quadratic term $\bf{Q}$ (\ie the Hessian), local invariance and selectivity directions can be analytically derived \cite{berkes2006analysis}, which correspond to eigenvectors of the least and most negative eigenvalues.
This method was also generalized onto multi-layer convolutional networks by numerically estimating the Hessian \cite{ngiam2010tiled}.
For multi-layer networks, characterization of the ``global'' invariance of single neurons can also be achieved by numerically searching for non-local (\ie farther to the optimal stimuli) solutions \cite{erhan2010understanding}.

@@

\section{Representation Characterization}

\begin{figure}
\begin{center}
\includegraphics[width=0.35\textwidth]{Figs/methods.pdf} 
\end{center}
\caption{{\bf Representation characterization methods.}
On the spherical (\ie constant energy) constraint of the $N$ dimensional stimulus space, similar to previous work, the first-order property (\ie optimal stimulus $\hat{\bf{x}}$, as shown in the figure's center) of a representation $\bf{r}$ is first iteratively searched (as the evolution of light gray eclipses shown in the figure) and analyzed.
However, unlike most previous work, the second-order properties (\ie invariant path $\lbrace \bf{x}^{+}_{\delta} \rbrace$ as dashed red curve and selectivity path $\lbrace \bf{x}^{-}_{\delta} \rbrace$ as dashed blue curve) at distances $0.1\pi$ to $0.5\pi$ from the optimal stimulus are also searched and analyzed.
Additionally, these methods are capable of characterizing both scalar and vector representations, where the ``fitnesses'' to be optimized correspond to Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1} and Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}, respectively.
In this example, the scalar representation (\ie response) of the given neuron is tuned to a $45^{\circ}$ Gabor filter and invariant to phase changes, but selective to orientation changes.
}
\label{fig:methods}
\end{figure}

@@

To study the representations inside an artificial neural network, methods adopted in this work include: (1) first-order (\ie linear) characterization, or search and analysis of optimal stimuli, and (2) second-order (\ie quadratic) characterization, or search and analysis of invariance and selectivity paths (\ie eigenvectors of the largest and smallest eigenvalues), which are generalized and improved from previous work.
$\bf{r}$ is defined as a scalar representation (\ie $\bf{r} \in \mathbb{R}$) when considering single neurons and a vector representation (\ie ${\bf{r}} \in \mathbb{R}^C$) when considering groups of neurons (with group size $C$).
Our methods support characterizing both scalar (using Eq.~\ref{eq:O1}, \ref{eq:I1}, \ref{eq:S1}, though Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2} can be used as well) and vector (using Eq.~\ref{eq:O2}, \ref{eq:I2}, \ref{eq:S2}) representations in a unified way.
The search phases defined via these equations will be introduced in the following subsections, and the analyses of search results using representation measures will be covered in Sec.~\ref{sec:exp}.

@@

While no specific knowledge of the network itself is assumed, we do restrict the space of stimuli considered with a constant energy constraint $\left\| \bf{x} \right\| = E$, where $E$ is set to $1$ in all mathematical derivations for simplicity, but set to the average energy of task-related images in our experiments.
Since neurons can often be modulated by stimulus contrasts, which is less interesting when considering pattern selectivities, limiting the stimulus space in this manner effectively avoids degenerate solutions that simply maximize stimulus contrasts, and reduces the range of possible stimuli to consider.

@@

In this work, the optimal stimulus is defined and numerically derived as 
\begin{align}
\hat{\bf{x}} &= \underset{\bf{x}} {\arg\max} f\left(\bf{x}\right) \label{eq:O1} \\
&= \underset{\bf{x}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}\right)-f\left(\tilde{\bf{x}}\right)\right\|} \right); \label{eq:O2}
\end{align}
subject to the spherical constraint $\left\| \bf{x} \right\| = 1$. 
As visualized in Fig.~\ref{fig:methods}, this process involves an iterative search on the non-convex fitness landscape, which in fact can be handled efficiently by most of the modern numerical solvers, especially when analytical derivatives (\ie backpropagation gradients) are available.
In this work, we adopted the CMA-ES solver \cite{hansen2001completely}, which does not rely on analytical gradients and instead stochastically estimates numerical derivatives (\ie inverse Hessian), and worked reasonably well in our experiments.
When considering a vector representation (using Eq.~\ref{eq:O2}), this process is equivalent to optimizing the response of an ``imaginary neuron'' tuned to a reference stimulus $\tilde{\bf{x}}$.

@@

With respect to the optimal stimulus $\hat{\bf{x}}$, the searches of invariance and selectivity paths, which consist of sets of invariant stimuli $\left\lbrace \bf{x}^{+}_{\delta} \right\rbrace$ and selective stimuli $\left\lbrace \bf{x}^{-}_{\delta} \right\rbrace$ respectively, are formulated as
\begin{align}
\bf{x}^{+}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\max} f\left(\bf{x}_{\delta}\right) \label{eq:I1} \\
&= \underset{\bf{x}_{\delta}}{\arg\max} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:I2} \\
\bf{x}^{-}_{\delta} &= \underset{\bf{x}_{\delta}}{\arg\min} f\left(\bf{x}_{\delta}\right) \label{eq:S1} \\
&= \underset{\bf{x}_{\delta}}{\arg\min} \left( e^{-\left\|f\left(\bf{x}_{\delta}\right)-f\left(\hat{\bf{x}}\right)\right\|} \right); \label{eq:S2}
\end{align}
in which $0 < \delta \le \frac{\pi}{2}$, subject to $\left\| \bf{x}_{\delta} \right\| = 1$ and $\langle \bf{x}_{\delta} , \hat{\bf{x}} \rangle = \cos\left(\delta\right)$. 
As visualized in Fig.~\ref{fig:methods}, the invariance path characterizes the $N$ dimensional curve that leads toward (or ideally, maintains at) a fitness as high as possible while moving away from the optimal stimulus, and the selectivity path characterizes the curve that leads toward fitness as low as possible while moving away.
The search process is implemented as multiple runs of maximization and minimization on discretized $\delta \in \left\lbrace 0.1\pi, 0.2\pi, 0.3\pi, 0.4\pi, 0.5\pi\right\rbrace$ as the circular distance constraints shown in Fig.~\ref{fig:methods}, where each run is initialized with the result from the previous run (and the $0.1\pi$ run directly with optimal stimulus $\hat{\bf{x}}$) to increase the path continuity and searching speed\footnote{\expdiff}.
In this work, the distance constraint $\delta$ only goes up to $\frac{\pi}{2}$, since on the $N$ dimensional sphere, stimuli that fall in $\frac{\pi}{2} < \delta \le \pi$ are simply ``negatives'' of those in $0 < \delta \le \frac{\pi}{2}$, which are of less uniqueness and interest in general. 
The same process can also be performed with respect to a reference stimulus $\tilde{\bf{x}}$, especially for vector representations, where the invariance and selectivity of the ``imaginary neuron'' are again characterized equivalently.

@@

For our experiments, a simplified variety of deep convolutional neural networks---the Spatio-Temporal Hierarchical Object Representation (STHOR) model \cite{pinto2009high, sthor}---{was} adopted.
Like other deep networks, it consists of the standard cascade of convolution, nonlinear activation, pooling layers, \etc (which together define a single ``level'' in this work). However, its convolution kernel weights are  randomly assigned rather than trained.
Though not matching the performances of deep networks trained on massive quantities of data \cite{krizhevsky2012imagenet}, it can in fact be surprisingly competitive for a variety of tasks, especially in the regime where large quantities of training samples are not readily available \cite{pinto2009high, cox2011beyond, viglarge}.
Most importantly, since it does not require training (except for the linear SVM on top of the representations extracted), a large number of networks with diverse hyperparameters can be rapidly generated and compared, which is particularly useful in this study where relative, instead of absolute task performances, are of primary interest. 
As one might be concerned that comparing randomly weighted networks will not tell much about representations, since they might as well be random, our results will suggest otherwise.

\newcommand{\expsettings}{In other words, 32 channels of filters in the top level's convolution layers, a simple setting with reasonable performance.
Following the terminology in recent work, shallow and deep neurons corresponded to \texttt{pool1} and \texttt{pool2} layers respectively.
Stimulus dimensionalities of the shallow and deep neurons {were} $N=121$ and $441$ (\ie spatially overlapping $11\times11$ and $21\times21$ receptive fields). % respectively
Other minor changes included: nonlinear activations all simplified to \emph{ReLU} \cite{krizhevsky2012imagenet} mode and normalizations all in subtractive mode.
Overall, the architectures were more similar to those in \cite{simonyan2014very}, except pooling operations can be \emph{average}, \emph{squared}, or \emph{max-like} in our case.
}

@@

In the experiments, 100 shallow (\ie one-level) networks and 100 deep (\ie two-level) networks {were} randomly generated (though all have 32 top-layer neurons\footnote{\expsettings}) and tested to see how their representations differ with networks' depths and affect networks' performances on face pair matching\footnote{\exptask} tasks against the LFW-a dataset \cite{LFWTech, wolf2011effective} as shown in Fig.~\ref{fig:perf}. %, using 10-fold cross validation on view 2.
The fitness landscapes of shallow and deep representations in our case are both non-convex, therefore there is no guarantee that optimality can be achieved through any optimization method.
However, by carefully adjusting the numerical solver and monitoring the quality of solutions, we argue that informative local optima can still be discovered and used to characterize the representations.
Specifically, all numerical optimizations {were} executed multiple times from different random initial stimuli, for increasing quality of the numerical solutions (of optimal stimuli and invariance and selectivity paths, with 2 runs executed and the better kept), and at the same time, for providing statistical samplings of the solution spaces (of encoding specificities and invariance and selectivity subspace properties, as detailed later).

@@

{\bf Optimal stimulus spectral complexity.} Figure \ref{fig:allrep}A and \ref{fig:allrep}E illustrate the trajectories of optimal stimulus searches of 32 top-layer neurons from the best performing shallow network and 32 from the best performing deep network respectively.
To further increase the searching speed, the initial stimulus {was} selected from 1000 ${1}/{f^{\alpha}}$ random stimuli with the best fitness, where $\alpha \in \left\lbrace -4,-3,-2,-1,0 \right\rbrace$, without overly sacrificing the nature of random initialization.
From the results, first, the response landscapes of deep neurons are noticeably more complex than those of shallow neurons, as a large fraction of searching trajectories for deep neurons actually go down then up before reaching the (local) optima, which also suggests the numerical solver can handle non-convexity reasonably well.
Second, the optimal stimuli of deep neurons are also relatively more complex (\ie consisting of more textural components) than those of shallow neurons (which are more uni-textured or Gabor-like).
To quantify such differences, the spectral complexity measure, which is estimated through the $L^{1}$ norm of the (2 dimensional) Fourier power spectrum of optimal stimulus, \ie $\left\| \mathcal{F}(\hat{\bf{x}}) \right\|_{1}$, is adopted, where a higher value suggests higher non-sparsity (\ie more spectral components required to represent the visual stimuli).

@@

{\bf Invariance subspace capacity.} To further characterize the invariance properties, multiple runs of invariance path searches {were} performed and the results are visualized in Figs.~\ref{fig:allrep}D and \ref{fig:allrep}H.
The diversity of the ``invariance plateau'' (\eg the central high fitness region in Fig.~\ref{fig:methods}) constitution of deep neurons is also more apparent than that of shallow neurons.
Compared to the invariance path potential, which is designed to characterize the best invariance path even if only one of such exists, we further adopted the invariance subspace capacity measure to estimates the ``dimensionality'' (\ie how diverse different paths can be) of the linear subspace formed by multiple path search results via the nuclear norm of concatenation of $n$ results $\left\|\left[{\bf{x}}_{\delta,1},\dots,{\bf{x}}_{\delta,n}\right]\right\|_{*}$ with $n=20$ and $\delta=0.1\pi$.

@@

\newcommand{\expregular}{While Equation \ref{eq:O2} describes the essence of our approach, it was not able to produce analyzable numerical results alone, due to the fact that deep representations can be highly invariant to noises that do not resemble natural images.
Implementation-wise, it was further regularized with the energy of Laplacian filtering $\left\| \mathcal{L} \left( \bf{x} \right) \right\|$ to enforce smoothness and similarity to natural images.
A similar concept was also adopted in~\cite{mahendran2014understanding}.
}  %, with regularization constant decided via optimizing withheld data.
